---
title: "Insert Title"
author: "Insert Names"
date: "Spring 2018"
output: html_document
urlcolor: blue
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Final Project

```{r}
#Load libraries
library(readr)
library(rvest)
library(dplyr)
library(ggplot2)
library(caret)

```


We only keep variables relevant to the state and that are not missing for any states.

```{r}
#Upload election integrity dataset
election_integrity_full <- read_csv("PEI US 2016 state-level (PEI_US_1.0) 16-12-2017.csv")

election_integrity <- election_integrity_full %>%
  select(stateabbr, PEIIndexi, PEIIndex_rank, PEItype, ratingstate, ratingstate_lci, ratingstate_hci, ratingcountry, ratingcountry_lci, ratingcountry_hci, lawsunfair2, favoredincumbent2, citizens2, lawsi, managed, votinginfo, fairofficials, legalelections, proceduresi, bfavored2, boundariesi, voteregi, oppprevent2, womenopp, minorityopp, leaderselect2, rallies2, partyregi, newspapers, tv2, fairaccess, faircoverage, mediai, donations, rich2, financei, violence2, fraudulent2, easy, choice, postal, expats, internet, rigged2, machinesaccurate, recordssecure, waited2, multiple2, regrestrictive2, popularwill, timely, votingi)

```


```{r}
#Upload voter turnout dataset
voter_turnout <- read_csv("2016 November General Election - Turnout Rates.csv")
voter_turnout <- voter_turnout %>%
  select(State, "VEP_Highest_Office" = "X5", "VAP_Highest_Office" = "X6", "Total_Ballots_Counted_(Estimate)" = "Numerators", "Highest_Office" = "X8", "Voting_Eligible_Population_VEP" = "Denominators", "Voting_Age_Population_VAP" = "X10", "Total_Ineligible_Felons" = "X15", "Abbreviation" = "X17") %>%
  filter(!is.na(State) & State != "United States") %>%
  mutate(Voting_Eligible_Population_VEP = as.numeric(gsub(",", "", as.character(Voting_Eligible_Population_VEP))), Voting_Age_Population_VAP = as.numeric(gsub(",", "", as.character(Voting_Age_Population_VAP))), Total_Ineligible_Felons = as.numeric(gsub(",", "", as.character(Total_Ineligible_Felons)))) %>%
  mutate(Voters_Disenfranchised = Voting_Age_Population_VAP - Voting_Eligible_Population_VEP, Pct_Felon_Disenfranchised = Total_Ineligible_Felons / Voting_Age_Population_VAP) %>%
  select(-State)

  
```

```{r}
#Scrape election results dataset
url <- "https://en.wikipedia.org/wiki/United_States_presidential_election,_2016"
tables <- url %>%
  read_html() %>%
  html_nodes(css = "table") 
election_results <- html_table(tables[[38]], fill = TRUE)
names(election_results) <- c("State", "Voting_Type", "Clinton_Number", "Clinton_Percent", "Clinton_Electoral", "Trump_Number", "Trump_Percent", "a", "b", "c", "d", "e", "f", "g", "h", "i", "j", "k", "l", "m", "n", "o", "p", "Abbreviation", "q")
election_results <- election_results %>%
  select(State, Clinton_Percent, Trump_Percent, Abbreviation) %>%
  filter(State != "U.S. Total", State != "Nebraska, 1st", State != "Nebraska, 2nd", State != "Nebraska, 3rd", State != "Maine, 1st", State != "Maine, 2nd", Clinton_Percent != "%")  %>%
  select(-State)

election_results$Abbreviation[election_results$Abbreviation == "ME–a/l"] <- "ME"
election_results$Abbreviation[election_results$Abbreviation == "NE–a/l"] <- "NE"

election_results <- election_results %>% 
  mutate(Clinton_Percent = as.numeric(sub("%", "", as.character(Clinton_Percent))), Trump_Percent = as.numeric(sub("%", "", as.character(Trump_Percent)))) %>%
  mutate(Vote_Clinton = ifelse(Clinton_Percent - Trump_Percent > 0, "Yes", "No"))

```

note inner_join, left_join, and right_join all do the same thing because we are matching 50 states
```{r}
#Merge datasets
integrity_turnout <- inner_join(election_integrity, voter_turnout, by = c("stateabbr" = "Abbreviation"))
all <- inner_join(integrity_turnout, election_results, by = c("stateabbr" = "Abbreviation"))
```

```{r}
#CW: Experimenting with PCA
pca_test <- all %>%
  select(-stateabbr, -PEIIndexi, -PEIIndex_rank, -PEItype, -ratingstate, -ratingstate_hci, -ratingstate_lci, -ratingcountry, -ratingcountry_lci, -ratingcountry_hci, -lawsi, -proceduresi, -boundariesi, -voteregi, -partyregi, -mediai, -financei, -VEP_Highest_Office, -VAP_Highest_Office, -`Total_Ballots_Counted_(Estimate)`, -Highest_Office, -Voting_Eligible_Population_VEP, -Voting_Age_Population_VAP, -Total_Ineligible_Felons, -Voters_Disenfranchised, -Clinton_Percent, -Trump_Percent, -Vote_Clinton, -votingi, -Pct_Felon_Disenfranchised)


prin_comp <- prcomp(pca_test, scale. = T)
biplot(prin_comp, scale = 0)

std_dev <- prin_comp$sdev
pr_var <- std_dev^2
prop_varex <- pr_var/sum(pr_var)
plot(cumsum(prop_varex), xlab = "Principal Component",
              ylab = "Cumulative Proportion of Variance Explained",
              type = "b")

plot(prop_varex, xlab = "Principal Component",
             ylab = "Proportion of Variance Explained",
             type = "b")

cw_data <- data.frame(state = all$stateabbr, index = all$PEIIndexi, clinton = all$Vote_Clinton, pct_dis = all$Pct_Felon_Disenfranchised, prin_comp$x) 

ggplot(cw_data, aes(x= index, y=PC1)) +
  geom_point(aes(col = clinton)) 

ggplot(cw_data, aes(x= PC2, y=PC1)) +
  geom_point(aes(col = clinton)) 

#Set cross validation options
cv_opts <- trainControl(method = "repeatedcv", number = 10, repeats = 5)

#Set seed
set.seed(7)

#Build the model
log_model1 <- train(clinton ~ index,  data=cw_data, method="glm", family="binomial",trControl = cv_opts)
log_model2 <- train(clinton ~ index + pct_dis,  data=cw_data, method="glm", family="binomial",trControl = cv_opts)
log_model3 <- train(clinton ~ PC1 + PC2 + PC3 + PC4 + PC5 + PC6 + PC7 + PC8 + PC9 + PC10 + PC11 + PC12 + PC13 +PC14 + PC15 + PC16 + PC17 + PC18 + PC19 + PC20,  data=cw_data, method="glm", family="binomial",trControl = cv_opts)

log_model1
log_model2
log_model3


```

From the PCA, we saw that it was possible for the PEIIndex to predict the outcome of an election. We wanted to test this out on a real life case study, and what better election to use than the highly contentious 2016 presidential election! Democrats frequently argue that Republicans pass laws to disenfranchise voters more likely to vote Democrat, giving them a greater chance that they'll win more elections. In this next section we will try to use the results the 2016 presidential election to gauge how much weight, if any, this argument has.
```{r}
#Creating a dataset with all the variables from election_integrity that couldn't be used for PCA because of missing values.
plots <- left_join(election_integrity_full, all)
#Getting a sense of voter access
ggplot(plots, aes(x = Pct_Felon_Disenfranchised, y = PEIIndexi, color = Vote_Clinton)) +
  geom_point()
ggplot(plots, aes(x = Vote_Clinton, y = Pct_Felon_Disenfranchised)) +
  geom_boxplot()
```

```{r}
ggplot(plots, aes(x = Vote_Clinton, y = boundariesi)) +
  geom_boxplot()
```

```{r}
ggplot(plots, aes(x = Vote_Clinton, y = citizens2)) +
  geom_boxplot()
```

```{r}
#Are there more illegal voters for Democrats?
ggplot(plots, aes(x = Vote_Clinton, y = ineligible2)) +
  geom_boxplot()
```


```{r}
#shiny
library(shiny)
library(plotly)
library(ggplot2)

all_quant <- all %>%
  select(stateabbr, Vote_Clinton, PEIIndexi, lawsi, proceduresi, boundariesi, voteregi, partyregi, mediai, financei, votingi, Pct_Felon_Disenfranchised) 

all_qual <- all %>%
  select(stateabbr, Vote_Clinton, lawsunfair2, favoredincumbent2, citizens2, managed, votinginfo, fairofficials, legalelections, bfavored2, oppprevent2, womenopp, minorityopp, leaderselect2, rallies2, newspapers, tv2, fairaccess, faircoverage, donations, rich2, violence2, fraudulent2, easy, choice, postal, expats, internet, rigged2, machinesaccurate, recordssecure, waited2, multiple2, regrestrictive2, popularwill, timely)

ui <- navbarPage(h2("Let's Look at Some Relationships!"),
  tabPanel("Quantitative Variables",
    plotlyOutput("quant"),
  
  hr(),
  
  fluidRow(
    column(5,
      selectInput("x", "X-Axis", names(all_quant)[3:11]),
      selectInput("y", "Y-Axis", names(all_quant)[3:11], names(all_quant)[4])
    ),
    column(2,
      checkboxGroupInput("candidate", "Candidate Chosen", 
                         choices = list("Clinton" = "Yes", "Trump" = "No"),
                         selected = unique(all_quant$Vote_Clinton))
            ),
    column(5,
      selectInput("state", "What state do you want to see?",
                  choices = unique(all_quant$stateabbr))
           )
  ),
  
  fluidRow(
    h3("Variable Key")
  )
),

  tabPanel("Categorical Variables",
    plotlyOutput("qual"),
      
  hr(),
  
  fluidRow(
    column(5,
      selectInput("x2", "X-Axis", names(all_qual)[3:34]),
      selectInput("y2", "Y-Axis", names(all_qual)[3:34], names(all_qual)[4])
    ),
    column(2,
      checkboxGroupInput("candidate2", "Candidate Chosen", 
                         choices = list("Clinton" = "Yes", "Trump" = "No"),
                         selected = unique(all_qual$Vote_Clinton))
            ),
    column(5,
      selectInput("state2", "What state do you want to see?",
                  choices = unique(all_qual$stateabbr))
           )
  ),
  
  fluidRow(
    h3("Variable Key"),
  p(tags$b("Note:"), "For all these variables, experts were asked to rate the various states on a scale of 1-5 based on various statements, with 1 being strongly agree and 5 being strongly disagree, which is why we called them categorical. However, all the experts' ratings were averaged, allowing them to be plotted as quantitative.")
  )
  )
)

server <- function(input, output) {

  output$quant <- renderPlotly({
    req(input$candidate)
    all_reactive <- filter(all_quant, Vote_Clinton %in% input$candidate)
    state_reactive <- filter(all_quant, stateabbr %in% input$state)
    ggplot(all_reactive, aes(color = Vote_Clinton, "State" = stateabbr)) +
      geom_point(aes_string(x = input$x, y = input$y)) +
      geom_text(data = state_reactive, mapping = aes_string(x = input$x, y = input$y), label = input$state) +
      scale_color_manual(values = c("red", "blue"), labels = c("Trump", "Clinton"), name = "") +
      coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))
  })
  
    output$qual <- renderPlotly({
    req(input$candidate2)
    reactive_all <- filter(all_qual, Vote_Clinton %in% input$candidate2)
    reactive_state <- filter(all_qual, stateabbr %in% input$state2)
    ggplot(reactive_all, aes(color = Vote_Clinton, "State" = stateabbr)) +
      geom_point(aes_string(x = input$x2, y = input$y2)) +
      geom_text(data = reactive_state, mapping = aes_string(x = input$x2, y = input$y2), label = input$state2) +
      scale_color_manual(values = c("red", "blue"), labels = c("Trump", "Clinton"), name = "") +
      coord_cartesian(xlim = c(0, 5), ylim = c(0, 5))
  })
}

shinyApp(ui, server)
```





