---
title: "How Fair Are Our Elections?"
author: "Chase Williamson and Tiffany Wang"
date: "Spring 2018"
runtime: shiny
urlcolor: blue
---

```{r setup, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Final Project

```{r}
#Load libraries
library(readr)
library(rvest)
library(dplyr)
library(ggplot2)
library(caret)
library(shiny)
library(plotly)
library(ggplot2)
library(ggthemes)
library(shinythemes)
library(ggmap)

```


We only keep variables relevant to the state and that are not missing for any states.

```{r}
#Upload election integrity dataset
election_integrity_full <- read_csv("PEI US 2016 state-level (PEI_US_1.0) 16-12-2017.csv")

election_integrity <- election_integrity_full %>%
  select(stateabbr, PEIIndexi, PEIIndex_rank, PEItype, ratingstate, ratingstate_lci, ratingstate_hci, ratingcountry, ratingcountry_lci, ratingcountry_hci, lawsunfair2, favoredincumbent2, citizens2, lawsi, managed, votinginfo, fairofficials, legalelections, proceduresi, bfavored2, boundariesi, voteregi, oppprevent2, womenopp, minorityopp, leaderselect2, rallies2, partyregi, newspapers, tv2, fairaccess, faircoverage, mediai, donations, rich2, financei, violence2, fraudulent2, easy, choice, postal, expats, internet, rigged2, machinesaccurate, recordssecure, waited2, multiple2, regrestrictive2, popularwill, timely, votingi)

```


```{r}
#Upload voter turnout dataset
voter_turnout <- read_csv("2016 November General Election - Turnout Rates.csv")
voter_turnout <- voter_turnout %>%
  select(State, "VEP_Highest_Office" = "X5", "VAP_Highest_Office" = "X6", "Total_Ballots_Counted_(Estimate)" = "Numerators", "Highest_Office" = "X8", "Voting_Eligible_Population_VEP" = "Denominators", "Voting_Age_Population_VAP" = "X10", "Total_Ineligible_Felons" = "X15", "Abbreviation" = "X17") %>%
  filter(!is.na(State) & State != "United States") %>%
  mutate(Voting_Eligible_Population_VEP = as.numeric(gsub(",", "", as.character(Voting_Eligible_Population_VEP))), Voting_Age_Population_VAP = as.numeric(gsub(",", "", as.character(Voting_Age_Population_VAP))), Total_Ineligible_Felons = as.numeric(gsub(",", "", as.character(Total_Ineligible_Felons)))) %>%
  mutate(Voters_Disenfranchised = Voting_Age_Population_VAP - Voting_Eligible_Population_VEP, Pct_Felon_Disenfranchised = Total_Ineligible_Felons / Voting_Age_Population_VAP) %>%
  select(-State)

  
```

```{r}
#Scrape election results dataset
url <- "https://en.wikipedia.org/wiki/United_States_presidential_election,_2016"
tables <- url %>%
  read_html() %>%
  html_nodes(css = "table") 
election_results <- html_table(tables[[38]], fill = TRUE)
names(election_results) <- c("State", "Voting_Type", "Clinton_Number", "Clinton_Percent", "Clinton_Electoral", "Trump_Number", "Trump_Percent", "a", "b", "c", "d", "e", "f", "g", "h", "i", "j", "k", "l", "m", "n", "o", "p", "Abbreviation", "q")
election_results <- election_results %>%
  select(State, Clinton_Percent, Trump_Percent, Abbreviation) %>%
  filter(State != "U.S. Total", State != "Nebraska, 1st", State != "Nebraska, 2nd", State != "Nebraska, 3rd", State != "Maine, 1st", State != "Maine, 2nd", Clinton_Percent != "%")  %>%
  select(-State)

election_results$Abbreviation[election_results$Abbreviation == "ME–a/l"] <- "ME"
election_results$Abbreviation[election_results$Abbreviation == "NE–a/l"] <- "NE"

election_results <- election_results %>% 
  mutate(Clinton_Percent = as.numeric(sub("%", "", as.character(Clinton_Percent))), Trump_Percent = as.numeric(sub("%", "", as.character(Trump_Percent)))) %>%
  mutate(Vote_Clinton = ifelse(Clinton_Percent - Trump_Percent > 0, "Clinton", "Trump"))

```

note inner_join, left_join, and right_join all do the same thing because we are matching 50 states + DC
```{r}
#Merge datasets
integrity_turnout <- inner_join(election_integrity, voter_turnout, by = c("stateabbr" = "Abbreviation"))
all <- inner_join(integrity_turnout, election_results, by = c("stateabbr" = "Abbreviation"))
```

```{r}
#Create a dataset for PCA. Note we only use variables on a scale from 0 to 5
pca_test <- all %>%
  select(-stateabbr, -PEIIndexi, -PEIIndex_rank, -PEItype, -ratingstate, -ratingstate_hci, -ratingstate_lci, -ratingcountry, -ratingcountry_lci, -ratingcountry_hci, -lawsi, -proceduresi, -boundariesi, -voteregi, -partyregi, -mediai, -financei, -VEP_Highest_Office, -VAP_Highest_Office, -`Total_Ballots_Counted_(Estimate)`, -Highest_Office, -Voting_Eligible_Population_VEP, -Voting_Age_Population_VAP, -Total_Ineligible_Felons, -Voters_Disenfranchised, -Clinton_Percent, -Trump_Percent, -Vote_Clinton, -votingi, -Pct_Felon_Disenfranchised)

#Run principle components
prin_comp <- prcomp(pca_test, scale. = T)

#Create variables necessary to explain amount of variance explained by various principal components
std_dev <- prin_comp$sdev
pr_var <- std_dev^2
prop_varex <- pr_var/sum(pr_var)

#Plot cumulative variance explained by principal components
plot(cumsum(prop_varex), xlab = "Principal Component",
              ylab = "Cumulative Proportion of Variance Explained",
              type = "b")

#Plot marginal variance explained by principal components
plot(prop_varex, xlab = "Principal Component",
             ylab = "Proportion of Variance Explained",
             type = "b")

#Create a data frame with state, index, whether the state voted for Trump or Clinton, and the principal components
all_with_pca <- data.frame(state = all$stateabbr, index = all$PEIIndexi, clinton = all$Vote_Clinton, prin_comp$x) 

#Plot PC1 against index variable
ggplot(all_with_pca, aes(x= index, y=PC1)) +
  geom_point(aes(col = clinton)) + 
  labs(x = "Election Integrity Index (Imputed)") +
  scale_color_manual(labels = c("Trump", "Clinton"), values = c("firebrick3", "royalblue2"), name = "Who Won the State?") +
  theme_classic() +
  theme(legend.position = c(.80,.85), legend.box.background = element_rect())

#Plot PC1 against PC2
ggplot(all_with_pca, aes(x= PC2, y=PC1, color = as.factor(clinton))) +
  geom_point() +
  scale_color_manual(labels = c("Trump", "Clinton"), values = c("firebrick3", "royalblue2"), name = "Who Won the State?") +
  theme_classic() +
  theme(legend.position = c(.17,.80), legend.box.background = element_rect())

#Set cross validation options
cv_opts <- trainControl(method = "cv", number = 51)

#Set seed
set.seed(10)

#Build the model only using the index variable
log_model1 <- train(clinton ~ index,  data=all_with_pca, method="glm", family="binomial",trControl = cv_opts)

#Build the model using first 6 principal components
log_model2 <- train(clinton ~ PC1 + PC2 + PC3 + PC4 + PC5,  data=all_with_pca, method="glm", family="binomial",trControl = cv_opts)

#Print out accuracy of models 1 and 2
log_model1
log_model2
```

From the PCA, we saw that it was possible for the PEIIndex to predict the outcome of an election. We wanted to test this out on a real life case study, and what better election to use than the highly contentious 2016 presidential election! Democrats frequently argue that Republicans pass laws to disenfranchise voters more likely to vote Democrat, giving them a greater chance that they'll win more elections. In this next section we will try to use the results the 2016 presidential election to gauge how much weight, if any, this argument has.
```{r}
#Creating a dataset with all the variables from election_integrity that couldn't be used for PCA because of missing values.
plots <- left_join(election_integrity_full, all)
#Getting a sense of voter access
ggplot(plots, aes(x = Pct_Felon_Disenfranchised, y = PEIIndexi, color = Vote_Clinton)) +
  geom_point() +
  labs(y = "Election Integrity Index (Imputed)", x = "Dienfranchised Felons (% of total eligible pop.)") +
  scale_color_manual(labels = c("Trump", "Clinton"), values = c("firebrick3", "royalblue2"), name = "Who Won the State?") +
  theme_classic() +
  theme(legend.position = c(.85,.80), legend.box.background = element_rect())

ggplot(plots, aes(x = Vote_Clinton, y = Pct_Felon_Disenfranchised)) +
  geom_boxplot()+
  labs(x = "Who won the state?", y = "Dienfranchised Felons (% of total eligible pop.)") +
  theme_classic()
```

```{r}
ggplot(plots, aes(x = Vote_Clinton, y = boundariesi)) + 
  geom_boxplot() +
  labs(x = "Who won the state?", y = "Boundaries Integrity Index (Imputed)") +
  theme_classic()
```

```{r}
ggplot(plots, aes(x = Vote_Clinton, y = citizens)) +
  geom_boxplot() +
  labs(x = "Who won the state?", y = "How Much Did the State Restrict Citizens' Rights?") +
  theme_classic()
```

```{r}
#Are there more illegal voters for Democrats?
ggplot(plots, aes(x = Vote_Clinton, y = ineligible)) +
  geom_boxplot() +
  labs(x = "Who won the state?", y = "How Likely Was It That Ineligible Voters Were Registered?") +
  theme_classic()
```

```{r}
#Overall conclusion of the experts?
ggplot(plots, aes(x = Vote_Clinton, y = PEIIndexi)) +
  geom_boxplot() +
  labs(x = "Who won the state?", y = "PEI Index Value") +
  theme_classic()
```

```{r}
#Shiny app


#Nice Labels
x_label <- data.frame(var = c("PEIIndexi", "lawsi", "proceduresi", "boundariesi", "voteregi", "partyregi", "mediai", "financei", "votingi", "Pct_Felon_Disenfranchised"), names = c("Overall Election Integrity Index (Imputed)", "Legal Integrity Index (Imputed)", "Procedures Integrity Index (Imputed)", "Boundaries Integrity Index (Imputed)", "Voter Registration Integrity Index (Imputed)", "Party Registration Integrity Index (Imputed)", "Media Coverage Integrity Index (Imputed)", "Campaign Finance Integrity Index (Imputed)", "Voting Integrity Index (Imputed)", "Dienfranchised Felons (% of total eligible pop.)"))
                      
y_label <- data.frame(var = c("PEIIndexi", "lawsi", "proceduresi", "boundariesi", "voteregi", "partyregi", "mediai", "financei", "votingi", "Pct_Felon_Disenfranchised"), names = c("Overall Election Integrity Index (Imputed)", "Legal Integrity Index (Imputed)", "Procedures Integrity Index (Imputed)", "Boundaries Integrity Index (Imputed)", "Voter Registration Integrity Index (Imputed)", "Party Registration Integrity Index (Imputed)", "Media Coverage Integrity Index (Imputed)", "Campaign Finance Integrity Index (Imputed)", "Voting Integrity Index (Imputed)", "Dienfranchised Felons (% of total eligible pop.)"))

#Loading data for longitudes and latitudes of states for map
states <- map_data("state")
election_integrity_full <- election_integrity_full %>%
  mutate(state = tolower(state))
map <- inner_join(election_integrity_full, states, by = c("state" = "region"))
map <- inner_join(map, election_results, by = c("stateabbr" = "Abbreviation"))


all_quant <- all %>%
  select(stateabbr, Vote_Clinton, PEIIndexi, lawsi, proceduresi, boundariesi, voteregi, partyregi, mediai, financei, votingi, Pct_Felon_Disenfranchised) 

all_qual <- map %>%
  select(stateabbr, Vote_Clinton, lawsunfair, favoredincumbent, citizens, managed, votinginfo, fairofficials, legalelections, bfavored, oppprevent, womenopp, minorityopp, leaderselect, rallies, newspapers, tv, fairaccess, faircoverage, donations, rich, violence, fraudulent, easy, choice, postal, expats, internet, rigged, machinesaccurate, recordssecure, waited, multiple, regrestrictive, popularwill, timely, long, lat, group)


##### UI Side ######
ui <- navbarPage("Looking Deeper into Election Integrity",
  theme = shinythemes::shinytheme("sandstone"),
  tabsetPanel(type = "tabs",
  tabPanel("Quantitative Variables",
    fluidRow(
      p("In order to calculate the overall election integrity index, The Electoral Integrity Project created a few subcategories that aggregated ratings the various experts gave for some topics. Below, you can explore the relationships between the subcategory scores and the overall index.")
    ),
    plotlyOutput("quant"),
  
  hr(),
  fluidRow(
    column(6,
      selectInput("x", "X-Axis", 
                 c("Overall Election Integrity Index (Imputed)" = "PEIIndexi", 
                  "Legal Integrity Index (Imputed)" = "lawsi", 
                  "Procedures Integrity Index (Imputed)" = "proceduresi", 
                  "Boundaries Integrity Index (Imputed)" = "boundariesi",
                  "Voter Registration Integrity Index (Imputed)" = "voteregi", 
                  "Party Registration Integrity Index (Imputed)" = "partyregi", 
                  "Media Coverage Integrity Index (Imputed)" = "mediai", 
                  "Campaign Finance Integrity Index (Imputed)" = "financei", 
                  "Voting Integrity Index (Imputed)" = "votingi")),
    
      selectInput(inputId = "y", "Y-Axis",
                  c("Overall Election Integrity Index (Imputed)" = "PEIIndexi", 
                  "Legal Integrity Index (Imputed)" = "lawsi", 
                  "Procedures Integrity Index (Imputed)" = "proceduresi", 
                  "Boundaries Integrity Index (Imputed)" = "boundariesi",
                  "Voter Registration Integrity Index (Imputed)" = "voteregi", 
                  "Party Registration Integrity Index (Imputed)" = "partyregi", 
                  "Media Coverage Integrity Index (Imputed)" = "mediai", 
                  "Campaign Finance Integrity Index (Imputed)" = "financei", 
                  "Voting Integrity Index (Imputed)" = "votingi"))
    ),
    column(2,
      checkboxGroupInput("candidate", "Candidate Chosen", 
                         choices = list("Clinton", "Trump"),
                         selected = unique(all_quant$Vote_Clinton))
            ),
    column(4,
      selectInput("state", "What state do you want to see?",
                  choices = unique(all_quant$stateabbr))
           )
  )
  ),

  tabPanel("Categorical Variables",
    plotOutput("qual"),
      
  hr(),
  
  fluidRow(
    column(5,
      selectInput("var", "What Variable Do You Want to Look At?", 
                  c("Laws are unfair" = "lawsunfair", "Incumbent is favored" = "favoredincumbent", "Citizens prevented from voting" = "citizens", "Elections well managed" = "managed", "Voting information widely available" = "votinginfo", "Officials were fair" = "fairofficials", "Elections followed the law" = "legalelections", "Boundaries favored incumbents" = "bfavored", "Opponent prevented from running" = "oppprevent", "Equal opportunities for women" = "womenopp", "Equal opportunities for minorities" = "minorityopp", "Top party leaders chose candidates" = "leaderselect", "Some candidates restricted from rallies" = "rallies", "Balanced newspaper coverage" = "newspapers", "TV favored governing party" = "tv", "Fair access to advertising" = "fairaccess", "Fair coverage of election" = "faircoverage", "Equal access to donations" = "donations", "Rich people buy elections" = "rich", "Some voters threatened at polls" = "violence", "Some fraudulent votes case" = "fraudulent", "Voting process was easy" = "easy", "Genuine choice at ballot box" = "choice", "Postal ballots available" = "postal", "Citizens abroad could vote" = "expats", "Internet voting available" = "internet", "Election was rigged" = "rigged", "Voting machines were accurate" = "machinesaccurate", "Voting records are secure" = "recordssecure", "People waited more than 30 min in line to vote" = "waited", "Some people voted more than once" = "multiple", "Registration deadlines too restrictive" = "regrestrictive", "Outcome reflected popular will" = "popularwill", "Ballots counted in timely fashion" = "timely"))
    ),
    
    column(2,
      checkboxGroupInput("candidate2", "Candidate Chosen", 
                         choices = list("Clinton", "Trump"),
                         selected = unique(all_qual$Vote_Clinton))
            ),
    
    column(5,
           p("Click the button to update the candidate!"),
           actionButton("action", "Update"))
  ),
  
  fluidRow(
  p(tags$b("Note:"), "For all these variables, experts were asked to rate the various states on a scale of 1-5 based on various statements, with 1 being strongly disagree and 5 being strongly agree, which is why we called them categorical. However, all the experts' ratings were averaged, allowing them to be plotted as quantitative.")
  )
  )
))

##### Server Side #####
server <- function(input, output) {
  
  #Label names
  xlab_var_name <- reactive({
    filter(x_label, var == input$x) %>%
    select(names) #.$names
  })
  
  ylab_var_name <- reactive({
    y_label %>%
    filter(var == input$y) %>%
    select(names) #.$names
  })
  
  output$quant <- renderPlotly({
    req(input$candidate)
    all_reactive <- filter(all_quant, Vote_Clinton %in% input$candidate)
    state_reactive <- filter(all_quant, stateabbr %in% input$state)
    ggplot(all_reactive, aes(color = Vote_Clinton, "State" = stateabbr)) +
      geom_point(alpha =.80, aes_string(x = input$x, y = input$y)) +
      geom_text(data = state_reactive, mapping = aes_string(x = input$x, y = input$y), label = input$state) +
      scale_color_manual(values = c("royalblue2", "firebrick3"), breaks = c("Trump", "Clinton"), name = "") +
      theme_classic() +
      labs(x = as.character(xlab_var_name()[1,1]), y = as.character(ylab_var_name()[1,1])) +
      theme(legend.position = c(.80,.85), legend.box.background = element_rect()) + 
      coord_cartesian(xlim = c(0, 100), ylim = c(0, 100))
  })
  
    output$qual <- renderPlot({
    qual_reactive <- eventReactive(input$action,
                            {select(all_qual,
                                    stateabbr, long, lat, Vote_Clinton, group, input$var)
                              filter(all_qual, Vote_Clinton == input$candidate2)
      })
    ggplot(data = qual_reactive()) + 
    geom_polygon(aes_string(x = "long", 
                     y = "lat", 
                     group = "group", 
                     fill = input$var), 
                     color = "white") + 
  scale_fill_gradient(high = "#E32F12", low = "#FBF0EE") +
  labs(fill = "Score") +
  coord_map() + 
  theme_void()
  })
}

shinyApp(ui, server)
```





